---
title: "Intermediate Concepts in R"
author: "Isaac Baca"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

---

# Introduction:

This project covers intermediate concepts in R.

---

# Table of contents
1. [Conditionals and Control Flow](#one)

      1.a [Relational Operators](#one-one)
    
      1.b [Logical Operators](#one-two)
      
      1.c [Conditional Statements](#one-three)
    
2. [Loops](#two)

      2.a [While loop](#two-one)
    
      2.b [For loop](#two-two)
    

3. [Functions](#three)

      3.a [Introduction to Functions](#three-one)
    
      3.b [Writing Functions](#three-two)

4. [The apply family](#four)

      4.a [lapply](#four-one)
    
      4.b [sapply](#four-two)
    
      4.c [vapply](#four-three)

5. [Utilities](#five)

      5.a [Useful Functions](#five-one)
    
      5.b [Regular Expressions](#five-two)
    
      5.c [Times & Dates](#five-three)


---

<a name="one"><a/>

## Conditionals and Control Flow

---

<a name="one-one"><a/>

### Relational Operators

---

#### The most basic form of comparison is equality.

Comparison of logicals

```{r echo=TRUE, message=FALSE, warning=FALSE}
TRUE == FALSE 
```


---

Comparison of numerics

```{r echo=TRUE, message=FALSE}
-6 * 14 != 17 - 101.
```


---

Comparison of character strings

```{r echo=TRUE, message=FALSE}
"useR" == "user"
```


---

Compare a logical with a numeric

```{r echo=TRUE, message=FALSE}
TRUE == 1
```


---

#### Greater and less than

Comparison of numerics

```{r echo=TRUE, message=FALSE}
-6 * 5 + 2 >= -10 + 1
```


---

Comparison of character strings

```{r echo=TRUE, message=FALSE}
"raining" <= "raining dogs"
```


---

Comparison of logicals

```{r echo=TRUE, message=FALSE}
TRUE > FALSE 
```


---

#### Compare matrices

I create two vectors (each containing the number of views for each day of the week), then combine them into a matrix called `views`.

```{r echo=TRUE, message=FALSE}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
facebook <- c(17, 7, 5, 16, 8, 13, 14)

views <- matrix(c(linkedin, facebook), nrow = 2, byrow = TRUE)
views
```


---

I use the views matrix to return a logical matrix to determine which observations are equal to 13.

```{r echo=TRUE, message=FALSE}
views == 13
```


---

I use the views matrix to return a logical matrix to determine which observations are less than or equal to 14.

```{r echo=TRUE, message=FALSE}
views <= 14
```


---

<a name="one-two"><a/>

### Logical Operators

#### & and |

Using the `|` sign will return `TRUE` if either statement is true.

Using the `&` sign will return `TRUE` if both statements are true.

---

I create a variable called `last` that holds the last observation in the `linkedin` vector.

```{r echo=TRUE, message=FALSE}
linkedin <- c(16, 9, 13, 5, 2, 17, 14)
last <- tail(linkedin, 1)

last
```


---

I check if last is under 5 or above 10?

```{r echo=TRUE, message=FALSE}
last < 5 | last > 10
```


---

I check if last is between 15 (exclusive) and 20 (inclusive)

```{r echo=TRUE, message=FALSE}
last > 15 & last <= 20
```


---

I check when LinkedIn views exceeded 10 and Facebook views failed to reach 10 for a particular day

```{r echo=TRUE, message=FALSE}
linkedin > 10 & facebook < 10
```


---

I check when one or both of the LinkedIn and Facebook profiles were visited at least 12 times.

```{r echo=TRUE, message=FALSE}
linkedin >=12 | facebook >=12
```


---

I check when the views matrix is equal to a number between 11 and 14, excluding 11 and including 14

```{r echo=TRUE, message=FALSE}
views > 11 & views <=14
```


---

<a name="one-three"><a/>

### Conditional Statements

#### The if statement

I set `medium` equal to "LinkedIn".

I write an if statement that prints out "Showing LinkedIn information" if the medium variable equals "LinkedIn".

```{r echo=TRUE, message=FALSE}
medium = "LinkedIn"

if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
}
```


---

I set `num_views` equal to 14.

I write an if statement that prints out "You are popular!" if `num_views` exceeds 14.

```{r echo=TRUE, message=FALSE}
num_views = 14

if(num_views > 15) {
  print("You are popular!")
}
```

`num_views` does not exceed 14.  There is no "else" statement, therefore there is no output.

---

#### Add an else statement

I add an else statement such that "Unknown medium" gets printed out to the console when the if-condition on medium does not hold.

```{r echo=TRUE, message=FALSE}
if (medium == "LinkedIn") {
  print("Showing LinkedIn information")
} else {
  print("Unknown medium")
}
```

The if-condition on medium holds, therefore the else statement does not execute.

---

I add an else statement such that R prints out "Try to be more visible!" when the if-condition on `num_views` is not met.

```{r echo=TRUE, message=FALSE}
if (num_views > 15) {
  print("You're popular!")
} else {
  print("Try to be more visible!")
}
```


---

#### The else if statement allows further customization of the control structure.

I add an else if statement such that "Your number of views is average" is printed if `num_views` is between 15 (inclusive) and 10 (exclusive).

```{r echo=TRUE, message=FALSE}
if (num_views > 15) {
  print("You're popular!")
} else if (num_views <= 15 & num_views > 10) {
print("Your number of views is average")
} else {
  print("Try to be more visible!")
}
```


---

I define two variables, `li` and `fb`, denoting the number of profile views the LinkedIn and Facebook profiles had on the last day of recordings.  I specify three possible conditions: 

1. If both li and fb are 15 or higher, set sms equal to double the sum of li and fb.

2. If both li and fb are strictly below 10, set sms equal to half the sum of li and fb.

3. In all other cases, set sms equal to li + fb.

```{r echo=TRUE, message=FALSE}
li <- 15
fb <- 9

if (li >=15 & fb >=15) {
  sms <- 2 * (li + fb)
} else if (li < 10 & fb < 10) {
  sms <- 0.5 * (li + fb)
} else {
  sms <- li + fb
}

print(sms)
```


---

<a name="two"><a/>

## Loops

---

<a name="two-one"><a/>

### While loop

---

#### Write a while loop

I code a while loop with the following characteristics:

- The condition of the while loop should check if speed is higher than 30.

- Inside the body of the while loop, print out "Slow down!".

- Inside the body of the while loop, decrease the speed by 7 units and assign this new value to speed again. This step is crucial; otherwise the while loop will never stop and your session will expire.


```{r echo=TRUE, message=FALSE}
# Initialize the speed variable
speed <- 64

# Code the while loop
while (speed > 30) {
  print("Slow down!")
  speed <- speed - 7
}

# Print out the speed variable
print(speed)
```


---

#### I add more conditionals to the loop

Check if speed is greater than 30.  

If the speed is greater than 48, I print "Slow down big time!", and decrease the speed by 11.

Otherwise, I print "Slow down!", and decrease the speed by 6.


```{r echo=TRUE, message=FALSE}
# Initialize the speed variable
speed <- 64

# Extend/adapt the while loop
while (speed > 30) {
  print(paste("Your speed is", speed))
  if (speed > 48) {
    print("Slow down big time!")
    speed = speed -11
  } else {
    print("Slow down!")
    speed = speed -6
  }
}
```


---

#### Break the loop

The break statement is a control statement. When R encounters it, the while loop is abandoned completely.

I adapt the while loop such that it is abandoned when the speed of the vehicle is greater than 80 (the speed variable has been initialized to 88)


```{r echo=TRUE, message=FALSE}
# Initialize the speed variable
speed = 88

while (speed > 30) {
  print(paste("Your speed is", speed))
  
  # Break the while loop when speed exceeds 80
  if (speed > 80) {
    break
  } 
  
  if (speed > 48) {
    print("Slow down big time!")
    speed = speed - 11
  } else {
    print("Slow down!")
    speed = speed - 6
  }
}
```


---

<a name="two-two"><a/>

### For loop

---

#### Loop over a vector

I write a for loop that iterates over all the elements of linkedin and prints out every element separately. 

There are two ways to do this:

- Loop version 1

- Loop version 2


```{r echo=TRUE, message=FALSE}
# The same linkedin vector from before
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# Loop version 1
for (views in linkedin) {
    print(views)
}


# Loop version 2
for(i in 1:length(linkedin)) {
    print(linkedin[i])
}
```


---

#### Loop over a list

I create a list called `nyc`

I look over the list using the two versions of loops

```{r echo=TRUE, message=FALSE}
# The nyc list
nyc = list(pop = 8405837, 
            boroughs = c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island"), 
            capital = FALSE)

# Loop version 1
for (info in nyc) {
    print(info)
}


# Loop version 2
for (i in 1:length(nyc)) {
    print(nyc[[i]])
}
```


---

#### Loop over a matrix

I create a matrix `ttt` that represents the status of a tic-tac-toe game.

```{r echo=TRUE, message=FALSE}
line1 = c("O", "NA", "X")
line2 = c("NA", "O", "O")
line3 = c("X", "NA", "X")

ttt = matrix(c(line1, line2, line3), nrow = 3, byrow = TRUE)
ttt
```

---

I create a nested for loop to go over the elements in `ttt`.

```{r echo=TRUE, message=FALSE}
for (i in 1:nrow(ttt)) {
  for (j in 1:ncol(ttt)) {
    print(paste("On row", i, "and column", j, "the board contains", ttt[i,j]))
  }
}
```


---

#### Using the if and else statements inside the for loop

I create a for loop that loops over the elements of the linkedin vector:

If the vector element's value exceeds 10, I print out "You're popular!".

If the vector element's value does not exceed 10, I print out "Be more visible!"


```{r echo=TRUE, message=FALSE}
# Same linkedin vector as before
linkedin <- c(16, 9, 13, 5, 2, 17, 14)

# the for loop with conditionals
for (i in 1:length(linkedin)) {
  if (linkedin[i] > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  print(linkedin[i])
}
```


---

#### Using the break and next statements:

- The break statement abandons the active loop: the remaining code in the loop is skipped and the loop is not iterated over anymore.

- The next statement skips the remainder of the code in the loop, but continues the iteration.

```{r echo=TRUE, message=FALSE}
# Same linkedin vector as before
linkedin = c(16, 9, 13, 5, 2, 17, 14)

for (li in linkedin) {
  if (li > 10) {
    print("You're popular!")
  } else {
    print("Be more visible!")
  }
  
  # Add if statement with break
  if(li > 16) {
    print("This is ridiculous, I'm outta here!")
    break
  }
  
  # Add if statement with next
  if(li < 5) {
    print("This is too embarrassing!")
    next
  }
  
  print(li)
}
```


---



```{r echo=TRUE, message=FALSE}

```


---

<a name="three"><a/>

## Functions

---

<a name="three-one"><a/>

### Introduction to Functions

All the relevant details such as a description, usage, and arguments can be found in the documentation. 

- To consult the documentation on the `mean()` function, use the following R command: `help(mean)`.

- To see the arguments of the sample() function, use `args(mean)`.

---

<a name="three-two"><a/>

### Writing Functions

#### Creating a function in R basically is the assignment of a function object to a variable

I create a function `pow_two()`, that takes one argument and returns that number squared.

I then call this function with `12` as input. 

```{r echo=TRUE, message=FALSE}
pow_two = function(x) {
    x^2
}

pow_two(12)
```


---

I create a function `sum_abs()`, that takes two arguments and returns the sum of the absolute values of both arguments

I then call the function `sum_abs()` with arguments `-2` and `3`

```{r echo=TRUE, message=FALSE}
sum_abs = function(x, y) {
    sum(abs(x), abs(y))
}

sum_abs(-2, 3)
```


---

#### Functions can be created with or without default values

I update the function `pow_two()`, adding an optional argument named `print_info` that is TRUE by default

I wrap an if construct around the `print()` function: this function should only be executed if `print_info` is TRUE

```{r echo=TRUE, message=FALSE}
pow_two <- function(x, print_info = TRUE) {
  y <- x ^ 2
  if(print_info == TRUE) {
    print(paste(x, "to the power two equals", y))
  } else {
    return(y)
  }
}


```


---

I call the previously defined function `pow_two()` without specifying the `print_info` argument.

```{r echo=TRUE, message=FALSE}
pow_two(12)
```


---

I call the previously defined function `pow_two()` with `print_info = FALSE`.

```{r echo=TRUE, message=FALSE}
pow_two(12,  print_info = FALSE)
```


---

<a name="four"><a/>

## The apply family

#### `lapply()` 

 - Apply a function over a list or vector
 
 - Always returns a list
 
#### `sapply()` 

 - Apply a function over a list or vector
 
 - Try to simplify the list to an array
 
#### `vapply()` 

 - Apply a function over a list or vector
 
 - Explicitly specify output format

---

<a name="four-one"><a/>

### lapply

`lapply(X, FUN, ...)` takes a vector or list `X`, and applies the function `FUN` to each of its members. If `FUN` requires additional arguments, you pass them after you've specified `X` and `FUN` `(...)`. 

The output of `lapply()` is a list, the same length as `X`, where each element is the result of applying `FUN` on the corresponding element of `X`.

- To convert this list to a vector, simply wrap the `lappy()` function inside the `unlist()` function. 

---

#### Useing lapply with a built-in R function

The `strsplit()` function takes two arguments:

1. A vector of strings.

2. An argument that tells the function where to split the strings.

I create a vector, `pioneers` of famous statisticians and the year they were born, then use `strsplit()` to split the strings in `pioneers` on the `:` sign.

```{r echo=TRUE, message=FALSE}
# The vector pioneers
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year
split_math <- strsplit(pioneers, split = ":")

# view the structure of split_math
str(split_math)
```

The result, `split_math` is a list of 4 character vectors: the first vector element represents the name, the second element the birth year.

---

I use `lapply()` to convert the character vectors in `split_math` to lowercase letters. 

Essentially, this will apply `tolower()` on each of the elements in `split_math`. 

I assign the result, which is a list, to a new variable `split_low` and view the structure. 

```{r echo=TRUE, message=FALSE}
split_low = lapply(split_math, tolower)

str(split_low)
```


---

#### Creating a new function to use inside lapply.

I create a new function `select_first()` that takes a vector as input and returns the first element of this vector.

```{r echo=TRUE, message=FALSE}
select_first <- function(x) {
  x[1]
}
```


---

I apply `select_first()` over the elements of `split_low` with `lapply()`.
```{r echo=TRUE, message=FALSE}
lapply(split_low, select_first)
```


---

I create a new function `select_second()` that takes a vector as input and returns the second element of this vector.

```{r echo=TRUE, message=FALSE}
select_second = function(x) {
  x[2]
}
```


---

I apply the `select_second()` function over `split_low`.

```{r echo=TRUE, message=FALSE}
lapply(split_low, select_second)
```


---

#### lapply and anonymous functions

Functions without names are called *anonymous functions*.

Anonymous functions can be used inside `lapply()`.

Instead of using the name of the functions that I created, I use the functions directly. 

```{r echo=TRUE, message=FALSE}
lapply(split_low, function(x) { x[1] })

lapply(split_low, function(x) { x[2] })
```


---

#### Using lapply with additional arguments

`lapply()` provides a way to handle functions that require more than one argument.

I define a function `select_el()` that takes a vector as its first argument, and an index as its second argument. It returns the vector's element at the specified index.

```{r echo=TRUE, message=FALSE}
select_el <- function(x, index) {
  x[index]
}
```


---

I use `lapply()` twice to call `select_el()` over all elements in `split_low`: 

Once with the index equal to 1 and a second time with the index equal to 2.

```{r echo=TRUE, message=FALSE}
lapply(split_low, select_el, 1)
lapply(split_low, select_el, 2)
```


---

<a name="four-two"><a/>

### sapply

`sapply(X, FUN, ..., USE.NAMES = TRUE)` is short for *simplify apply*.  It is used similar to how `lapply()` is used.  

The output of `lapply()` is a named vector that contains the same information as the vector obtained from using unlist and lapply together, `unlist(lapply())`. 

The names can be omitted by setting the `USE.NAMES` argument to `FALSE`.

There are cases in which the function you want to apply does not always return a vector of the same length at all times.  In this case `sapply()` will return a list. 

---

#### I create a list named `temp`, that contains temperature measurements for 7 days. `Temp` is a list of length 7, where each element is a vector of length 5, representing 5 measurements on a given day.

```{r echo=TRUE, message=FALSE}
temp = list(c(3, 7,  9,  6, -1), 
            c(6,  9, 12, 13,  5), 
            c(4,  8,  3, -1, -3), 
            c(1,  4,  7,  2, -2),
            c(5,  7,  9,  4,  2), 
            c(-3,  5,  8,  9,  4), 
            c(3 , 6,  9,  4,  1))

temp
```


---

I use `lapply()` to calculate the minimum of the temperature measurements for every day.

```{r echo=TRUE, message=FALSE}
lapply(temp, min)
```


---

I do the same thing, but with sapply.

```{r echo=TRUE, message=FALSE}
sapply(temp, min)
```


Instead of a list, I have a vector.

---

#### sapply with a self-defined function. 

I define a function `extremes_avg()` that takes a vector of temperatures and calculates the average of the minimum and maximum temperatures of the vector.


```{r echo=TRUE, message=FALSE}
extremes_avg <- function(x) {
  ( min(x) + max(x) ) / 2
}
```


---

I use the self-defined function inside lapply.

```{r echo=TRUE, message=FALSE}
lapply(temp, extremes_avg)
```


---

I use the self-defined function inside sapply.

```{r echo=TRUE, message=FALSE}
sapply(temp, extremes_avg)
```


---

I define a function `extremes()` that takes a vector of numerical values and returns a vector containing the minimum and maximum values of a given vector, with the names "min" and "max", respectively.

```{r echo=TRUE, message=FALSE}
extremes <- function(x) {
  c(min = min(x), max = max(x))
}
```


---

I apply this function over the vector `temp` using lapply.

```{r echo=TRUE, message=FALSE}
lapply(temp, extremes)
```


---

I apply this function over the vector `temp` using sapply.

```{r echo=TRUE, message=FALSE}
sapply(temp, extremes)
```


---

#### There are cases in which the function you want to apply does not always return a vector of the same length at all times.  In this case `sapply()` will return a list.

I define a function `below_zero()` that takes a vector of numerical values and returns a vector that only contains the values that are strictly below zero.

```{r echo=TRUE, message=FALSE}
below_zero = function(x) {
  return(x[x < 0])
}
```


---

I apply `below_zero()` over temp using `lapply()` and store the result in `freezing_l`.

```{r echo=TRUE, message=FALSE}
freezing_l = lapply(temp, below_zero)

freezing_l
```


---

I apply `below_zero()` over temp using `sapply()` and store the result in `freezing_s`.

```{r echo=TRUE, message=FALSE}
freezing_s = sapply(temp, below_zero)

freezing_s
```


---

I compare `freezing_s` to `freezing_l` using the `identical()` function.

```{r echo=TRUE, message=FALSE}
identical(freezing_l, freezing_s)
```


---

<a name="four-three"><a/>

### vapply 

`vapply(X, FUN, FUN.VALUE, ..., USE.NAMES = TRUE)` is used similar to how `lapply()` and `sapply()` is used.

The output of `lapply()` has to be specified using the `FUN.VALUE` argument.

`Vapply()` is safer than `sapply()` if you want to simplify the result that `lapply()` generates because if the specified output is not possible, it generates an error. 

---

I create a function `basics()` that takes a vector, and returns a named vector of length 3, containing the minimum, mean and maximum value of the vector respectively.

```{r echo=TRUE, message=FALSE}
basics <- function(x) {
  c(min = min(x), mean = mean(x), max = max(x))
}
```


---

I apply the function `basics()` over the list of temperatures, `temp`, using `vapply()`.

I use `numeric(3)` to specify the `FUN.VALUE` argument because the expected output for each vector is a vector of length three.

```{r echo=TRUE, message=FALSE}
vapply(temp, basics, numeric(3))
```


---

I change the function `basics()` to include the median.

```{r echo=TRUE, message=FALSE}
basics <- function(x) {
  c(min = min(x), mean = mean(x), median = median(x), max = max(x))
}
```


---

I use `vapply()` setting the `FUN.VALUE` argument to `numeric(4)`.

```{r echo=TRUE, message=FALSE}
vapply(temp, basics, numeric(4))
```


---

<a name="five"><a/>

## Utilities

---

<a name="five-one"><a/>

### Useful Functions

---

#### Mathematical utilities

Some useful math functions include:

- `abs()`: Calculate the absolute value.

- `sum()`: Calculate the sum of all the values in a data structure.

- `mean()`: Calculate the arithmetic mean.

- `round()`: Round the values to 0 decimal places by default. Try out `?round` in the console for variations of `round()` and ways to change the number of digits to round to.

---

I calculate the sum of the absolute rounded values of the vector `numbers`.

```{r echo=TRUE, message=FALSE}
numbers = c(1.9, -2.6, 4.0, -9.5, -3.4, 7.3)

print(sum(round(abs(numbers))))
```


---

I define a vector of numbers `vec1` and a second vector, `vec2` which reverses the order of `vec1`

I then find the mean of the absolute values of the combined vectors.

```{r echo=TRUE, message=FALSE}
vec1 <- c(1.5, 2.5, 8.4, 3.7, 6.3)
vec2 <- rev(vec1)

mean(c(abs(vec1), abs(vec2)))
```


---

#### Useful data utilities:

- `seq()`: Generate sequences, by specifying the from, to, and by arguments.

- `rep()`: Replicate elements of vectors and lists.

- `sort()`: Sort a vector in ascending order. Works on numerics, but also on character strings and logicals.

- `rev()`: Reverse the elements in a data structures for which reversal is defined.

- `str()`: Display the structure of any R object.

- `append()`: Merge vectors or lists.

- `is.*()`: Check for the class of an R object.

- `as.*()`: Convert an R object from one class to another.

- `unlist()`: Flatten (possibly embedded) lists to produce a vector.

---

I create two lists, one called `linkedin` and one called `facebook`

```{r echo=TRUE, message=FALSE}
linkedin <- list(16, 9, 13, 5, 2, 17, 14)
facebook <- list(17, 7, 5, 16, 8, 13, 14)
```


---

I view the lists

```{r echo=TRUE, message=FALSE}
linkedin
facebook
```


---

I convert these lists to vectors using `unlist()`

```{r echo=TRUE, message=FALSE}
li_vec = unlist(linkedin)
fb_vec = unlist(facebook)
```


---

I view the vectors

```{r echo=TRUE, message=FALSE}
li_vec
fb_vec
```


---

I append `fb_vec` to the `li_vec` and call the new vector `social_vec`.

```{r echo=TRUE, message=FALSE}
social_vec = append(li_vec, fb_vec)

social_vec
```


---

I sort `social_vec` from high to low.

```{r echo=TRUE, message=FALSE}
sort(social_vec, decreasing = TRUE)
```


---

Using the function `seq()`, I create a sequence that ranges from 1 to 500 in increments of 3. I name the vector `seq1`. 

```{r echo=TRUE, message=FALSE}
seq1 = seq(1, 500, by = 3)

seq1
```


---

I create a sequence that ranges from 1200 to 900 in increments of -7.

```{r echo=TRUE, message=FALSE}
seq2 = seq(1200, 900, by = -7)

seq2
```


---

I calculate the sum of the two sequences.

```{r echo=TRUE, message=FALSE}
print(sum(c(seq1, seq2)))
```


---

<a name="five-two"><a/>

### Utilities

In their most basic form, regular expressions can be used to see whether a pattern exists inside a character string or a vector of character strings. For this purpose, you can use:

---

#### grepl & grep

`grepl()`, returns `TRUE` when a pattern is found in the corresponding character string.

`grep()`, returns a vector of indices of the character strings that contains the pattern.

Both functions need a `pattern` and an `x` argument, where `pattern` is the regular expression you want to match for, and the `x` argument is the character vector from which matches should be sought.

---

I first define a vector `emails`.

```{r echo=TRUE, message=FALSE}
emails <- c("john.doe@ivyleague.edu", "education@world.gov", "dalai.lama@peace.org",
            "invalid.edu", "quant@bigdatacollege.edu", "cookie.monster@sesame.tv")
```


---

Use `grepl()` to generate a vector of logicals that indicates whether these email addresses contain "edu".

```{r echo=TRUE, message=FALSE}
grepl("edu", emails)
```


---

I do the same thing with `grep()`, but this time I save the resulting indexes in a variable `hits`.

```{r echo=TRUE, message=FALSE}
hits = grep("edu", emails)

hits
```


---

I use the variable `hits` to select from the `emails` vector only the emails that contain "edu".

```{r echo=TRUE, message=FALSE}
emails[hits]
```


---

I use the variable `hits` to select from the `emails` vector only the emails that *do not* contain "edu".

```{r echo=TRUE, message=FALSE}
emails[-hits]
```


---

<a name="five-three"><a/>

### Times and Dates

Dates are represented by `Date` objects, while times are represented by `POSIXct` objects. Under the hood, however, these dates and times are simple numerical values. `Date` objects store the number of days since the 1st of January in 1970. `POSIXct` objects on the other hand, store the number of seconds since the 1st of January in 1970.

Of course, it's also possible to create dates and times before 1970; the corresponding numerical values are simply negative in this case.

---

I ask R for the current date, and store the result in a variable `today`.

```{r echo=TRUE, message=FALSE}
today = Sys.Date()

today
```


---

To see what today looks like under the hood, I call `unclass()` on it.

```{r echo=TRUE, message=FALSE}
print(paste(unclass(today), "days have passed since Jan 1st, 1970"))
```


---

I ask R for the current time, and store the result in a variable `now`.

```{r echo=TRUE, message=FALSE}
now = Sys.time()

now
```


---

To see what now looks like under the hood, I call `unclass()` on it.

```{r echo=TRUE, message=FALSE}
print(paste(unclass(now), "seconds have passed since Jan 1st, 1970"))
```


---

#### Creating and formating dates

To create a Date object from a simple character string in R, you can use the `as.Date()` function. The character string has to obey a format that can be defined using a set of symbols (the examples correspond to 13 January, 1982):

- %Y: 4-digit year (1982)

- %y: 2-digit year (82)

- %m: 2-digit month (01)

- %d: 2-digit day of the month (13)

- %A: weekday (Wednesday)

- %a: abbreviated weekday (Wed)

- %B: month (January)

- %b: abbreviated month (Jan)

The following R commands will all create the same Date object for the 13th day in January of 1982:

- `as.Date("1982-01-13")`
- `as.Date("Jan-13-82", format = "%b-%d-%y")`
- `as.Date("13 January, 1982", format = "%d %B, %Y")`

Notice that the first line here did not need a format argument, because by default R matches your character string to the formats `"%Y-%m-%d"` or `"%Y/%m/%d"`.

In addition to creating dates, you can also convert dates to character strings that use a different date notation. For this, you use the `format()` function:

`today <- Sys.Date()`
`format(Sys.Date(), format = "%d %B, %Y")`
`format(Sys.Date(), format = "Today is a %A!")`

---

I define three character strings representing dates

```{r echo=TRUE, message=FALSE}
str1 <- "May 23, '96"
str2 <- "2012-03-15"
str3 <- "30/January/2006"
```


---

I convert them to dates using `as.Date()`, and assign them to `date1`, `date2`, and `date3` respectively.

```{r echo=TRUE, message=FALSE}
date1 <- as.Date(str1, format = "%b %d, '%y")
date2 = as.Date(str2)
date3 = as.Date(str3, format = "%d/%B/%Y")
```


---

From the first date, I use `format()` to extract the weekday.

```{r echo=TRUE, message=FALSE}
format(date1, "%A")
```


---

From the second date, I use `format()` to extract the day of the month.

```{r echo=TRUE, message=FALSE}
format(date2, "%d")
```


---

From the third date, I use `format()` to extract the abbreviated month and the 4-digit year, separated by a space.

```{r echo=TRUE, message=FALSE}
format(date3, "%b %Y")
```


---

#### Creating and formating times.

Similar to working with dates, `as.POSIXct()` can be used to convert from a character string to a `POSIXct` object, and `format()` to convert from a `POSIXct` object to a character string. Again, there are a wide variety of symbols:

- %H: hours as a decimal number (00-23)

- %I: hours as a decimal number (01-12)

- %M: minutes as a decimal number

- %S: seconds as a decimal number

- %T: shorthand notation for the typical format %H:%M:%S

- %p: AM/PM indicator

For a full list of conversion symbols, consult the `strptime` documentation:

Again, `as.POSIXct()` uses a default format to match character strings. In this case, it's `%Y-%m-%d %H:%M:%S`.\

---

I define two strings that represent time stamps, `str1` and `str2`.

```{r echo=TRUE, message=FALSE}
str1 <- "May 23, '96 hours:23 minutes:01 seconds:45"
str2 <- "2012-3-12 14:23:08"
```


---

I convert `str1` and `str2` to `POSIXct` objects called `time1 and `time2`.

```{r echo=TRUE, message=FALSE}
time1 <- as.POSIXct(str1, format = "%B %d, '%y hours:%H minutes:%M seconds:%S")
time2 = as.POSIXct(str2)
```


---

I use `format()` to create a string from `time1` containing only the minutes.

```{r echo=TRUE, message=FALSE}
format(time1, format = "%M")
```


---

From `time2`, I extract the hours and minutes as "hours:minutes AM/PM".

```{r echo=TRUE, message=FALSE}
format(time2, format = "%I:%M %p")
```


---

#### Calculations with Times

Calculations using `POSIXct` objects are completely analogous to those using `Date` objects.

I define two vectors: 

- `logout` which contains user logout times for a website.

- `login` which contains user longin times for a website.

```{r echo=TRUE, message=FALSE}
logout = c("2023-04-20 10:56:29 UTC", "2023-04-25 09:14:52 UTC", "2023-04-25 12:35:48 UTC", "2023-04-25 13:17:22 UTC", "2023-04-27 22:08:47 UTC")

login = c("2023-04-20 10:18:04 UTC", "2023-04-25 09:14:18 UTC", "2023-04-25 12:21:51 UTC", "2023-04-25 12:37:24 UTC", "2023-04-27 21:37:55 UTC")
```


---

I check the data type of `logout` and `login`.

```{r echo=TRUE, message=FALSE}
class(logout)
class(login)
```


These are characters and they need to be `POSIXct`.

---

I change the data type to `POSIXct` and check the data type again. 

```{r echo=TRUE, message=FALSE}
logout = as.POSIXct(logout)
login = as.POSIXct(login)

class(logout)
class(login)
```


They are now in the correct data type. 

---

I calculate the difference between the two vectors `logout` and `login`, i.e. the time the user was online in each independent session. The results are stored in a variable `time_online`.

```{r echo=TRUE, message=FALSE}
time_online = logout - login

time_online
```


---

I calculate the total time online.  

Be default R will return the time in seconds.  I divide the result by 60 to convert seconds to minutes. 

```{r echo=TRUE, message=FALSE}
sum(time_online)

print(paste(7426/60, "min"))
```


---

I calculate the average time online per session. 

Be default R will return the time in seconds.  I divide the result by 60 to convert seconds to minutes. 

```{r echo=TRUE, message=FALSE}
mean(time_online)

print(paste(1485.2/60, "min"))
```


---

I create two vectors:

- A vector `astro`, which contains character strings representing the dates on which the four astronomical seasons start.

- A vector `meteo`, which contains character strings representing the dates on which the four meteorological seasons start.

```{r echo=TRUE, message=FALSE}
astro = c("20-Mar-2015", "25-Jun-2015", "23-Sep-2015", "22-Dec-2015")
meteo = c("March 1, 15", "June 1, 15", "September 1, 15", "December 1, 15")
```


---

I check the data type of these two vectors

```{r echo=TRUE, message=FALSE}
class(astro)
class(meteo)
```


---

I change the data type to `Date`, specifying the format.

```{r echo=TRUE, message=FALSE}
astro_dates = as.Date(astro, format = "%d-%b-%Y")
meteo_dates = as.Date(meteo, format = "%B %d, %y") 
```


---

I calculate the maximum absolute difference between the astronomical and the meteorological beginnings of a season.

```{r echo=TRUE, message=FALSE}
max(abs(meteo_dates - astro_dates))
```


